Cross-validation of regression models: mixed model examples (edited v3)

### 2024-07-26 scratch/JSS-submission/Fox-Monette-cv-package.Rmd
further edit to the JSS paper

Validating research by replication on independently collected data is a common scientific norm. Emulating this process in a single study by data-division is less common [see, e.g., @Barnard:1974]: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data. Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, the results can vary substantially based on the random division of the data, particularly in smaller samples. See @Harrell:2015 [Sec. 5.3] for cogent remarks about data-division and cross-validation.

@@

The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical models. The principal function in the packages, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
```{r cv-methods}
library("cv", quietly = TRUE)
methods("cv")
```
 The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.

@@

```{r Auto}
data("Auto", package = "ISLR2")
summary(Auto)
```
With the exception of \code{origin} (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see \code{help("Auto",} \code{package = "ISLR2")}.

@@

* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} (the default) selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).

@@

[^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100. With the exception of the last column, the output is self-explanatory. The last column shows which methods have average timings that are statistically distinguishable. Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the patient reader to redo this computation with the default \code{times = 100} repetitions.

@@

To generate the four data sets, we first constructed a template data set from a population with a between-patient effect of dosage of $1.0$ and a within-patient effect of $-0.5$. We then applied four different multipliers to the between-patient variation from the regression line of patient centroids. The resulting data sets, shown in Figure \ref{fig:plot1}, differ in the between-patient variance of patient centroids from a common between-patient regression line, exhibiting a progression from low to high variance around the common regression line. Each patient has the same relative configuration of dosages and symptoms in each of the four data sets. To help visualize the structure of the data, estimated 50% Gaussian concentration ellipses are shown for each patient.

@@

```{r plot-fits-combined,echo=FALSE}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient standard deviation ratios. The top row shows summaries of the within-patient data using estimated 50% concentration ellipses, along with the patient centroids."
levs3 <- levels(Data$panel2)
levs3 <- sub(':fixed','', levs3)
levs3 <- sub(':BLUPs',' ', levs3)
Data$panel3 <- factor(sub(':fixed','', Data$panel2,), levels = levs3)
Datap$panel3 <- factor(sub(':BLUPs',' ', Data$panel2,), levels = levs3)

@@

We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, \code{~ 1}  to the larger \code{~ 1 + x} model when the between-cluster variation is relatively small, but that cluster-based cross-validation will prefer the latter to the former. We will discover that case-based cross-validation prefers the \code{~ 1 + x} model to the \code{~ 1} model for the '5 / 2.5' data set, but cluster-based cross-validation prefers the latter model to the former. The situation is entirely reversed with the '8 / 2.5' data set.

@@

```{r Auto-explore-scatterplot-matrix}
#| out.width = "95%",
#| fig.height = 5.5,
#| fig.align = 'center',
#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data."
Auto$cylinders <- factor(Auto$cylinders,
                         labels = c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels = c("AMER", "EUR", "JAP"))
rownames(Auto) <- make.names(Auto$name, unique = TRUE)
Auto$name <- NULL

@@

* \code{age}, in years;
* \code{econ.cond.national} and \code{econ.cond.household}, the respondent's ratings of the state of the economy, on 1 to 5 scales;
* \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales;
* \code{Europe}, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism";
* \code{political.knowledge}, knowledge of the parties' positions on European integration, with scores from 0 to 3;
* \code{gender}, \code{"female"} or \code{"male"}.

@@

rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
rs_obj$accuracy <- map_dbl(rs_obj$results, function(x) mean(x$correct))
1 - summary(rs_obj$accuracy) 
```
The response variable in this example, \code{Attrition}, is binary. The function \code{holdout_results()} must be supplied by the user, and "\code{accuracy}" is the complement of the Bayes-rule classification error rate.

@@

The \pkg{caret} package supports generalized linear models fit by \code{glm()} and can straightforwardly perform LOO CV:
```{r rsample-example-using-caret, cache=TRUE}
library("ggplot2", warn.conflicts = FALSE)
library("caret", quietly = TRUE, warn.conflicts = FALSE)
train(x = attrition[, c("JobSatisfaction", "Gender", "MonthlyIncome")],
      y = attrition$Attrition, method = "glm",
      trControl = trainControl(method = "LOOCV"))
```
where "\code{Accuracy}", again, is the complement of the misclassification rate.

@@

The weighted-least-squares (WLS) estimator of $\boldsymbol{\beta}$ is [see, e.g., @Fox:2016, Sec. 12.2.2] [^WLS]
$$
\mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^\top \mathbf{W} \mathbf{X} \right)^{-1} 
  \mathbf{X}^\top \mathbf{W} \mathbf{y}
$$

@@

To compute cross-validated fitted values when the folds contain more than one case, we make use of the Woodbury matrix identity [@Hager:1989],
$$
\left(\mathbf{A}_{m \times m} + \mathbf{U}_{m \times k} 
\mathbf{C}_{k \times k} \mathbf{V}_{k \times m} \right)^{-1} = \mathbf{A}^{-1} - \mathbf{A}^{-1}\mathbf{U} \left(\mathbf{C}^{-1} + 
\mathbf{VA}^{-1}\mathbf{U} \right)^{-1} \mathbf{VA}^{-1}
$$
where $\mathbf{A}$ is a nonsingular order-$n$ matrix. We apply this result by letting
\begin{align*}
	\mathbf{A} &= \mathbf{X}^\top \mathbf{W} \mathbf{X} \\
	\mathbf{U} &= \mathbf{X}_\mathbf{j}^\top \\
	\mathbf{V} &= - \mathbf{X}_\mathbf{j} \\
	\mathbf{C} &= \mathbf{W}_\mathbf{j} \\
\end{align*}
where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jm})^\top$ represents the vector of indices for the cases in the $j$th fold, $j = 1, \ldots, k$. The negative sign in $\mathbf{V} = - \mathbf{X}_\mathbf{j}$ reflects the *removal*, rather than addition, of the cases in $\mathbf{j}$.

>>>>

Validating research by replication on independently collected data is a common scientific norm. Emulating this process in a single study by data-division is less common [see, e.g., @Barnard:1974]: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit (in the language of machine leaning, "train") a statistical model; and then the second part is used to assess the adequacy of ("validate") the model fit to the first part of the data. Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, the results can vary substantially based on the random division of the data, particularly in smaller samples. See @Harrell:2015 [Sec. 5.3] for cogent remarks about data-division and cross-validation.

@@

The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical models. The principal function in the package, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
```{r cv-methods}
library("cv", quietly = TRUE)
methods("cv")
```
 The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.

@@

```{r Auto}
data("Auto", package = "ISLR2")
summary(Auto)
```
With the exception of \code{origin} (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: For details see \code{help("Auto",} \code{package = "ISLR2")}.

@@

* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} (the default) selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than directly recomputing the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).

@@

[^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100. With the exception of the last column, the output is self-explanatory. The last column shows which methods have average timings that are statistically distinguishable (as indicated by different letters). Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the patient reader to redo this computation with the default \code{times = 100} repetitions.

@@

To generate the four data sets, we first constructed a template data set from a population with a between-patient effect of dosage of $1.0$ and a within-patient effect of $-0.5$. We then applied four different multipliers to the between-patient standard deviation from the regression line of patient centroids. The resulting data sets, shown in Figure \ref{fig:plot1}, differ in the between-patient variation of patient centroids from a common between-patient regression line, exhibiting a progression from low to high variation around the common regression line. Each patient has the same relative configuration of dosages and symptoms in each of the four data sets. To help visualize the structure of the data, estimated 50% Gaussian concentration ellipses are shown for each patient.

@@

```{r plot-fits-combined,echo=FALSE}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient standard-deviation ratios. The top row shows summaries of the within-patient data using estimated 50% concentration ellipses, along with the patient centroids."
levs3 <- levels(Data$panel2)
levs3 <- sub(':fixed','', levs3)
levs3 <- sub(':BLUPs',' ', levs3)
Data$panel3 <- factor(sub(':fixed','', Data$panel2,), levels = levs3)
Datap$panel3 <- factor(sub(':BLUPs',' ', Data$panel2,), levels = levs3)

@@

We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, \code{~ 1}  to the larger \code{~ 1 + x} model when the between-cluster variation is relatively small, but that cluster-based cross-validation will prefer the latter to the former. We will discover that case-based cross-validation prefers the \code{~ 1 + x} model to the \code{~ 1} model for the "5 / 2.5" data set, but cluster-based cross-validation prefers the latter model to the former. The situation is entirely reversed with the "8 / 2.5" data set.

@@

```{r Auto-explore-scatterplot-matrix}
#| out.width = "95%",
#| fig.height = 5.5,
#| fig.align = 'center',
#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data."
Auto$cylinders <- factor(Auto$cylinders,
                         labels = c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels = c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique = TRUE)
Auto$name <- NULL

@@

* \code{age}, in years.
* \code{econ.cond.national} and \code{econ.cond.household}, the respondent's ratings of the state of the economy, on 1 to 5 scales.
* \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales.
* \code{Europe}, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism."
* \code{political.knowledge}, knowledge of the parties' positions on European integration, with scores from 0 to 3.
* \code{gender}, \code{"female"} or \code{"male"}.

@@

rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
rs_obj$accuracy <- map_dbl(rs_obj$results, function(x) mean(x$correct))
1 - summary(rs_obj$accuracy) 
```
The response variable in this example, \code{Attrition}, is binary. The function \code{holdout_results()} must be supplied by the user, and mean "\code{accuracy}" is the complement of the Bayes-rule classification error rate.

@@

The \pkg{caret} package supports generalized linear models fit by \code{glm()} and can straightforwardly perform LOO CV:
```{r rsample-example-using-caret, cache=TRUE}
library("ggplot2", warn.conflicts = FALSE)
library("caret", quietly = TRUE, warn.conflicts = FALSE)
train(x = attrition[, c("JobSatisfaction", "Gender", "MonthlyIncome")],
      y = attrition$Attrition, method = "glm",
      trControl = trainControl(method = "LOOCV"))
```
where "\code{Accuracy}," again, is the complement of the misclassification rate.

@@

The weighted-least-squares ("WLS") estimator of $\boldsymbol{\beta}$ is [see, e.g., @Fox:2016, Sec. 12.2.2] [^WLS]
$$
\mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^\top \mathbf{W} \mathbf{X} \right)^{-1} 
  \mathbf{X}^\top \mathbf{W} \mathbf{y}
$$

@@

To compute cross-validated fitted values when the folds contain more than one case, we make use of the Woodbury matrix identity [@Hager:1989],
$$
\left(\mathbf{A}_{m \times m} + \mathbf{U}_{m \times k} 
\mathbf{C}_{k \times k} \mathbf{V}_{k \times m} \right)^{-1} = \mathbf{A}^{-1} - \mathbf{A}^{-1}\mathbf{U} \left(\mathbf{C}^{-1} + 
\mathbf{VA}^{-1}\mathbf{U} \right)^{-1} \mathbf{VA}^{-1}
$$
where $\mathbf{A}$ is a nonsingular order-$m$ matrix. We apply this result by letting
\begin{align*}
	\mathbf{A} &= \mathbf{X}^\top \mathbf{W} \mathbf{X} \\
	\mathbf{U} &= \mathbf{X}_\mathbf{j}^\top \\
	\mathbf{V} &= - \mathbf{X}_\mathbf{j} \\
	\mathbf{C} &= \mathbf{W}_\mathbf{j} \\
\end{align*}
where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jn_j})^\top$ represents the vector of indices for the $n_j \approx n/k$ cases in the $j$th fold, $j = 1, \ldots, k$. The negative sign in $\mathbf{V} = - \mathbf{X}_\mathbf{j}$ reflects the *removal*, rather than addition, of the cases in $\mathbf{j}$.
